Customizing Software by Direct Manipulation of Tabular Data

### 2020-05-21 paper.tex
related work

Most attempts at empowering end users to customize their software offer
a simplified version of programming. Some scripting languages
(AppleScript, Chickenfoot) have a friendly syntax that resembles natural
language. Visual programming tools (Mac Automator, Zapier) eliminate
text syntax entirely. Macro recorders (Applescript, Helena, WebVCR)
remove some of the initial programming burden by letting a user start
with concrete demonstrations.

@@

Some customization environments (Mac Automator, Zapier) forego text
syntax and enable the user to connect programs and construct automations
by dragging and dropping commands. These environments still do not
constitute direct manipulation, though: the objects being manipulated
are in the domain of programming, not the domain of the task at hand.

@@

This paper extends work reported in a workshop paper by Litt and Jackson
\citep{litt2020} which presented an earlier prototype version of the
Wildcard system. We build on their work in this paper by rearchitecting
the internal implementation of the system around the table adapter
abstraction, evaluating it on many more websites, and by more fully
characterizing the behavior and design of the system.

@@

\begin{itemize}
\item
  Buttons \citep{maclean1990}: the original DM customization tool
\item
  ScrAPIr: direct manipulation for APIs!
\item
  customization tools

  \begin{itemize}
  \item
    browser extensions
  \item
    Thresher
  \item
    Sifter
  \item
    customization research tools: Chickenfoot, Coscripter
  \item
    Wildcard, like Chickenfoot, wants to hide HTML from users. But we
    show a structured data view, whereas Chickenfoot shows nothing
  \item
    visual customization tools:

    \begin{itemize}
    \tightlist
    \item
      Mac Automator
    \item
      Zapier
    \item
      IFTTT
    \item
      These are weakly direct manipulation. But usually the thing you
      directly manipulate in these UIs is the \emph{script}: not the
      actual \emph{data} you want to operate on. VERY DIFFERENT.
    \end{itemize}
  \item
    desktop customization: Applescript, VBA, COM
  \end{itemize}
\end{itemize}

\hypertarget{visual-database-query-tools}{%
\subsection{Visual database query
tools}\label{visual-database-query-tools}}

\begin{itemize}
\tightlist
\item
  database GUIs:

  \begin{itemize}
  \tightlist
  \item
    Sieuferd \citep{bakke2016}
  \item
    Airtable \citep{2020a}
  \item
    other similar tools?
  \item
    Liu \& Jagadish: rules for a spreadsheet algebra for database
    queries \citep{liu2009}
  \item
    Shneiderman paper
  \item
    spreadsheets
  \end{itemize}
\end{itemize}

\hypertarget{other}{%
\subsection{Other}\label{other}}

\begin{itemize}
\tightlist
\item
  web scraping / wrapper induction
\item
  instrumental interaction, Scotty \citep{eagan2011}, Webstrates
\item
  browser dev tools
\item
  Self, Smalltalk
\item
  multiple representations
\item
  Semantic Web
\end{itemize}

>>>>

Most attempts at empowering end users to customize their software offer
a simplified version of programming. Some scripting languages
\citep{bolin2005, cook2007} have a friendly syntax that resembles
natural language. Some visual customization tools eliminate text syntax
entirely. Macro recorders \citep{cook2007, chasins2018, anupam2000}
remove some of the initial programming burden by letting a user start
with concrete demonstrations.

@@

Some customization environments (Automator for Mac, Shortcuts for iOS,
Zapier for web APIs) forego text syntax and enable the user to connect
programs and construct automations by dragging and dropping commands.
These environments still do not constitute direct manipulation, though:
the objects being manipulated are in the domain of programming, not the
domain of the task at hand.

@@

This paper extends work reported in a workshop paper by Litt and Jackson
\citep{litt2020} which presented an early prototype version of Wildcard.
We have substantially extended their work in this paper by creating the
table adapter abstraction and reimplementing the system around that
abstraction, evaluating the system more fully on many more websites, and
by characterizing the design of the system in more detail.

Table-driven customization relates to two broad areas of related work.
Our problem statement is related to software customization tools, and
our solution approach is related to spreadsheets and other direct
manipulation interfaces.

@@

Table-driven customization is most closely related to other tools that
aim to empower end users to customize software without doing any
programming.

This lineage goes back at least to the Buttons system by MacLean et al
\citep{maclean1990}, where Xerox Lisp users could share buttons that
performed various ``tailoring'' actions on the system. The authors
proposed the ``gentle slope'' idea which has greatly influenced our
approach to table-driven customization (as discussed in
Section~\ref{sec:dm}). The authors also point out the importance of a
``tailoring culture'' where people with different skillsets collaborate
to produce useful customizations; in their system, Lisp programmers
create buttons that others can use, modify, and rearrange. This division
of labor corresponds to our idea of semantic wrappers, where end user
customization is supported by programmer-created building blocks.

Some recent web customization tools aim to enable end users to modify
web interfaces without programming.

Sifter \citep{huynh2006} enables end users to sort and filter lists of
data based on web scraping, much like Wildcard's sorting features. The
main difference between the systems is that table-driven customization
has many other use cases besides sorting and filtering. Also, Sifter
involves end users in a semi-automated data extraction process, rather
than having programmers create wrappers. This provides coverage of more
websites, but at the expense of complicating the end user experience. We
might explore such techniques in Wildcard in the future, but we think
that it's valuable for end users to have a customization experience free
of web scraping details on supported websites. Finally Sifter implements
scraping across multiple pages, a valuable feature for sorting and
filtering which isn't present in Wildcard.

Thresher \citep{hogue2005} enables end users to create wrappers which
map website content to Semantic Web schemas, making customizations
available using the schema information. We share the general idea of
wrappers with Thresher, but we have chosen to map to a single generic
data type, rather than more specific schemas, which increases the range
of websites and data supported by the system.

There are many software customization tools that offer simplified
versions of programming for end users. Chickenfoot \citep{bolin2005} and
Coscripter \citep{leshed2008} offer friendly syntax for writing web
automation scripts; Applescript \citep{cook2007} attempts a similar goal
for desktop customization. There are visual programming environments for
customization which don't involve writing any text: as a few examples,
\href{https://support.apple.com/guide/automator/welcome/mac}{Automator}
for Mac and
\href{https://apps.apple.com/us/app/shortcuts/id915249334}{Shortcuts}
for iOS are modern options for customizing Apple products, and
\href{https://zapier.com/}{Zapier} enables users to connect different
web applications together visually. As mentioned previously, these tools
all require writing imperative programs in some form; we take a
fundamentally different approach in table-driven customization.

\hypertarget{direct-manipulation-programming-interfaces.}{%
\subsection{Direct manipulation programming
interfaces.}\label{direct-manipulation-programming-interfaces.}}

Another relevant area of literature is direct manipulation interfaces
for interacting with structured data. We take much inspiration from
these tools in our work, but apply them in a different domain:
customizing existing software applications, rather than interacting with
databases or constructing software from scratch.

The most closely related work is in systems that offer spreadsheet-like
querying of relational data, as proposed by Liu and Jagadish
\citep{liu2009}. SIEUFERD by Bakke and Karger \citep{bakke2016} is an
example; their paper presents a survey of various other similar tools.
Our work is particularly influenced by the authors' observation that
direct manipulation requires that the user manipulate the \emph{results}
of a database query rather than the query itself, and that the user must
see intermediate results at every step of constructing a computation.
SIEUFERD's interface supports a far wider range of possible queries than
Wildcard, but the basic ideas of the user interface are similar.

\href{https://airtable.com/}{Airtable} is another example of a modern
commercial product that offers spreadsheet-like interaction with a
relational database.

Our work is also inspired by the many projects which have explored using
spreadsheets as a foundation for building software applications,
including Object Spreadsheets \citep{mccutchen2016}, Quilt
\citep{benson2014}, Gneiss \citep{chang2014}, Marmite \citep{wong2007},
\href{https://www.glideapps.com/}{Glide}. These projects share the idea
that a spreadsheet is a convenient interface for editing the data
underlying a GUI application. We share that idea, but apply it to
software customization, rather than building software from scratch.

Another system in this space is ScrAPIr, by Alrashed et
al.~\citep{alrashed2020}, which enables end users to access backend web
APIs without programming. ScrAPIr shares a high level goal of end user
empowerment, and also shares the idea of wrappers, by creating a shared
library of wrappers around existing web APIs. However, unlike Wildcard,
ScrAPIr targets explicit APIs exposed by developers, and also doesn't
aim to extend the frontend interfaces of web applications.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

\emph{todo}
